Compositional Quantitative Resource Analysis of C Programs

### 2014-06-26 paper/main.tex
adjustments in the example descriptions

It is often necessary to relate certain invariants in the pre- and
postcondition when specifying a program.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving
triples in the logic, so the Coq theorem for the above example would
look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function $S_f$ body with this
induction hypothesis.  In this process it can be the case that the
induction hypothesis needs to be used with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

% As mentioned, our account of the logic in this article is slightly
% simplified compared to the Coq development to improve readability.
% % However, the described logic is very close to the implemented version.

% The main difference between the implemented logic and the logic
% described here is that the latter does not have an
% \emph{auxiliary state}.
% %
% Auxiliary state is a classic extension of Hoare logic (see for
% example~\cite{Nipkow02}).
%  It is used to share information
% between the pre- and postcondition of a triple.  In a logic without
% auxiliary state (or similar techniques) it is not possible to relate
% program states before and after a statement.  For example, you cannot
% specify that the function {\tt\ int\ twice\ ()\ $\{$i = i+i;$\}$\ }
% doubles the value of the variable $i$.  With an auxiliary variable $Z$
% it is possible specify this fact in Hoare logic using the triple $\htriple{i =
%   Z}{\code{twice()}}{ i = 2 \cdot Z}$.

% In the Coq implementation,
% assertions have the type $\State \times \mathit{Aux} \to \N \cup
% \{\infty\}$.  The auxiliary state is an arbitrary type in Coq and can
% be instantiated by the user.  Most of the rules of the logic remain
% unchanged in the presence of auxiliary state.  The
% only exception is the consequence rule \textsc{L:Conseq} that reads as
% follows.
% {  \footnotesize
% \begin{mathpar}
%   \RuleNolabel
%  {  \htriple{P'}{S}{R}
%  \\  
%  {\!\!\!\! \begin{array}{r}
% \forall \sigma \, a. P(\sigma,a) {\in} \N \! \implies  \! \exists a'.  
%      P(\sigma,a) {\geq} P'(\sigma,a') \hspace{3.3em} \\ \land (\forall \sigma'\, i.  R^i(\sigma',a') {\geq} Q^i(\sigma',a))
%  \end{array}}\!\!\!\!
%  }
%  { \htriple{P}{S}{Q}
%  }
% \end{mathpar}}
% This is the quantitative version of a consequence rule that has been
% introduced in the context of Hoare logic and is attributed to Martin
% Hofmann~\cite{Nipkow02}.  Its typical use case is when we apply the rule {\sc L:Call}
% to a recursive call.  In this case, the Hoare triple for the function
% call is proved by an assumption from the derivation context with
% a slightly different auxiliary state.  In the example derivation
% in~\pref{fig:bsearch_deriv} this different state is $Z - 1$.  Adapting
% the derivation hypothesis to prove the recursive call is enabled in
% our logic by the extended consequence rule introduced above.

% \paragraph{Stack Framing}

% A minor difference in the function application rule is that we only
% present the rule for function calls with a single argument and without
% \emph{framing of stack assertions}.  The latter is necessary in the
% code of a caller to carry over information on the local environment
% from the precondition of the function call to the postcondition of the
% function call.  This is a general problem in Hoare logic and stack
% framing is a well known solution.  For instance, we would like to prove
% something like $\htriple{y}{ f(x) }{y}$ where $y$ is a local variable.
% This is not possible with the rule \textsc{L:Call} but it is with the more
% general rule in the Coq implementation~\iffullversion{\cite{coqfiles}}{\cite{anoncoq}}.

@@

Our first example on~\pref{fig:xmplmax} uses the expressive power of
the logic to get a very precise runtime bound on a program searching
a maximal element in an array.  The metric used simply counts the
assignments performed by the program.  Hence, the resource consumption
is closely related to the number $A$ of times the test \lstinline{a[i] > m}
is true during the execution.
%
It is clear that $A+1$ is the count of ``maximums so far in the
array \lstinline{a}'':
$$
  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
$$
where we write $\# S$ for the cardinal of the set $S$.  This number is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But using the expressivity of our logic we can get a finer bound
that depends on the contents of the array before the program execution.

The only non-trivial part of this derivation is finding the loop
invariant that is used in the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program; it can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
is admissible in our system.

>>>>

When specifying a program, it is often necessary to relate certain
invariants in the pre- and postcondition.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before using the
logic to derive triples, so the Coq theorem for the above example
would look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function body $S_f$ with this
induction hypothesis.  In this process it can be the case that we
have to use the induction hypothesis with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

@@

Our first example on~\pref{fig:xmplmax} derives a precise runtime bound
on a program that searches a maximal element in an array.
The metric used simply counts the assignments performed by the program.
Hence, the resource consumption is closely related to the number $A$ of
times the test \lstinline{a[i] > m} is true during the execution.
%
It is clear that $A+1$ is the count of ``maximums so far in the
array \lstinline{a}'':
$$
  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
$$
where we write $\# S$ for the cardinal of the set $S$.  This number is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our logic it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

The non-trivial part of this derivation is finding the loop
invariant that for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program.  It can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
rule is admissible in our system.
